Notes on \\ Projected Knowledge Compilation \\ \today

### 2023-08-01 doc/pkc-notes.tex
Refinements

We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
\begin{description}
\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L = \emptyset\}$.

\item[$\negop(\phi, L)$:] Remove the negations of the literals from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
  Then $\negop(\phi, L) = \{ \negop(C, L) \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.

\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
$\simplify{\phi}{L} = \posop(\negop(\phi, L), L)$.

@@

When constructing POFs via a bottom-up process, we can perform a set of simplifications, as follows:

\begin{itemize}
\item
The use of negation can be optimized as follows:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Double negations: $\neg \neg \psi = \psi$
\end{itemize}

@@

Our goal is to implement a projecting knowledge compiler that produces a POF $\pkc(\phi)$ when given input formula $\phi$.
We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that simultaneously operates on a POF $\psi$ and a clausal formula $\phi$.  Projected knowledge compilation is then defined
as $\pkc(\phi) = \traverse(\kc(\phi), \bcp(\phi))$.

@@

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the product operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the sum operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yield subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi' \not = \tautology$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.

@@

@@

Figure \ref{fig:eg-proj}(A) shows the POG representation of the
example formula, derived from the dec-DNNF representation produced by
\dfour{}.  In this graph, product nodes are labeled with the operation $\pand$, and sum nodes are labeled with the operation $\por$, emphasizing that these operations satisfy their respective partitioning constraints.
We can see that \dfour{} used $y$ as the decision variable at the top level, creating sum node $\nodes_7$ having edges to
node $\nodep_5$ representing product operation $P(x_3, \obar{y})$ on the left and  node $\nodep_6$
representing product operation $P(x_1, x_2, y)$ on the right.

Figure \ref{fig:eg-proj}(B) shows the result of our algorithm, based on the following recursive steps.
Computing $\traverse(P(x_3, \obar{y}), \phi)$ yields the literal $x_3$, while
computing $\traverse(P(x_1, x_2, y), \phi)$ yields node $\nodep_{10}$ representing the product operation $P(x_1, x_2)$.
Computing $\ureduce{\phi}{y}$ yields the clausal formula $\phi'$ given as
\begin{center}
\begin{tabular}{ll}
\toprule
\makebox[10mm]{Clause} & \\
\midrule
\texttt{1} & \texttt{0} \\
\texttt{2} & \texttt{0} \\
 \texttt{3} & \texttt{0}\\
\bottomrule
\end{tabular}
\end{center}
and this is compiled to node $\nodep_{8}$ representing product operation $P(x_1, x_2, x_3)$.
Finally, the nodes are assembled using our DeMorgan's Law transformations as shown with root node $\nodes_{11}$.

>>>>

We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
\begin{description}
\item[$\posop(\phi, L)$:] Remove clauses satisfied by the literals in $L$: $\posop(\phi, L) = \{ C \in \phi \mid C \cap L = \emptyset\}$.

\item[$\negop(\phi, L)$:] Remove the negations of the literals in $L$ from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
  Then $\negop(\phi, L) = \{ \negop(C, L) \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.

\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals in $L$, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
$\simplify{\phi}{L} = \posop(\negop(\phi, L), L)$.

@@

When constructing a POF via a bottom-up process, we can perform simplifications as the formula is constructed, as follows:
\begin{itemize}
\item
The use of negation can be optimized as follows:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Double negations: $\neg \neg \psi = \psi$
\end{itemize}

@@

Our goal is to implement a projecting knowledge compiler that produces a POF $\pkc(\phi)$ when given input formula $\phi$.
We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that traverses a POF $\psi$ while maintaining its clausal formula $\phi$.  Projected knowledge compilation is then defined
as $\pkc(\phi) = \traverse(\kc(\phi), \bcp(\phi))$.

@@

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yield subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi' \not = \tautology$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
The two arguments to the top-level sum operation are disjoint, because the overlapping models have been subtracted from the first argument.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.

@@

The first two termination rules have some subtle difference.  For the
initial formulation of $\traverse$, the arguments will always satisfy
$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, the
pure literal reduction rule can eliminate some literals from $\phi$
that occur in $\psi$, and so the two sets of variables can become
incomparable.  When both rules are applicable, it is more efficient to use the first rule, since it avoids an invocation of $\kc$.

@@

Figure \ref{fig:eg-proj}(A) shows the POG representation of the
example formula, derived from the dec-DNNF representation produced by
\dfour{}.  In this graph, product nodes are labeled $P$, and sum nodes are labeled  $S$.
We can see that \dfour{} used $y$ as the decision variable at the top level, creating sum node $\nodes_7$ having edges to
node $\nodep_5$ representing product operation $P(x_3, \obar{y})$ on the left and  node $\nodep_6$
representing product operation $P(x_1, x_2, y)$ on the right.

Figure \ref{fig:eg-proj}(B) shows the result of our algorithm, based on the following recursive steps.
Computing $\traverse(P(x_3, \obar{y}), \phi)$ yields the literal $x_3$, while
computing $\traverse(P(x_1, x_2, y), \phi)$ yields node $\nodep_{10}$ representing the product operation $P(x_1, x_2)$.
Computing $\ureduce{\phi}{y}$ yields the clausal formula $\phi'$ given as
\begin{center}
\begin{tabular}{ll}
\toprule
\makebox[10mm]{Clause} & \\
\midrule
\texttt{1} & \texttt{0} \\
\texttt{2} & \texttt{0} \\
 \texttt{3} & \texttt{0}\\
\bottomrule
\end{tabular}
\end{center}
and this is compiled to node $\nodep_{8}$ representing product operation $P(x_1, x_2, x_3)$.
Finally, the nodes are assembled using our DeMorgan's Law transformations yielding a root edge to node $\nodes_{11}$.
